\\textsc{Mathematical Description}

### 2023-06-07 100-math-description/GR-sslcm-math-description.Rmd
Replace \rightarrow with \to
Much more readable

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \renewcommand*{\log}{\mathrm{log}\hspace{-0.20em}}
  - \renewcommand*{\max}{\mathrm{max}\hspace{-0.20em}}
  - \newcommand{\logit}{\mathrm{logit}\hspace{-0.20em}}
  - \newcommand{\ilogit}{\mathrm{logit}^{-1}\hspace{-0.20em}}
  - \newcommand{\MVNdist}{\mathcal{MVN}\hspace{-0.20em}}
  - \newcommand{\Ndist}{\mathcal{N}\hspace{-0.20em}}
  - \newcommand{\Bdist}{\mathcal{B}\hspace{-0.20em}}
  - \newcommand{\Mdist}{\mathcal{M}\hspace{-0.20em}}
---

@@

* For simplicity, we use the term "adult" to refer to any mature individual returning to spawn including age-3 "jacks".
* Life stage-specific states are defined by capital letters, e.g., parr ($P$), smolt ($M$), or spawners ($S$)
* Parameters (e.g., survival terms, non-survival transition probabilities, coefficients in relationships, etc.) are denoted by Greek symbols, e.g., $\phi$ for survival rates or $\psi$ for maturation rates.
* Subscripts denote the index of a quantity, e.g., $y$ denotes a specific year and $j$ a specific population (see Table \@ref(tab:index-table)).
* We use the convention $b$ to indicate "before event" and $a$ to indicate "after event" for a quantity that occurs at the same life stage. E.g., $M^{b}$ represents smolt abundance _before_ out-of-basin migration and $M^{a}$ represents smolt abundance _after_ out-of-basin migration. As such, superscripts are most often used syntactically, not mathematically. E.g., $M^{b}$ is not "$M$ raised to the power $b$"; if we must raise the quantity $M^{b}$ to the power $c$, it will be written $(M^{b})^c$.
* Survival terms to transition from state $A$ to state $B$ in year $y$ for population $j$ are denoted by $\phi^{A \rightarrow B}_{y,j}$.
* We denote the expected value of a stochastic process with a dot, e.g., $\dot{\phi}^{A \rightarrow B}_j$ -- this is the value expected in the absence of process variability.
* In some cases, we specify calculations performed on vectors rather than scalars (i.e., a scalar being a single element from a vector or higher-dimensional array). Vectors are denoted via boldface, and the dimension of the vector is denoted using 1:$n$ syntax. E.g., $\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}$ represents a vector of survival terms from state $A$ to state $B$ in year $y$, where each element of the vector represents the value for each population.
* Mathematical operations performed on two vectors $\boldsymbol{C}$ and $\boldsymbol{D}$ are performed in an element-wise fashion, and the resulting vector will have identical dimensions to both $\boldsymbol{C}$ and $\boldsymbol{D}$.
* In some cases, a calculation is performed differently for a subset of the possible index values. For example, to denote that $k$ should take on only the values of 1 and 3 (not 2), we would write, $k \in [1,3]$.
* $\log\left(x\right)$ represents the natural logarithm of $x$ and it is implied that the constraint $x > 0$ is satisfied.
* $\logit\left(p\right)$ represents the log odds of $p$ (i.e., $\log\left[p/(1-p)\right])$) and it is implied that the constraint $0 < p < 1$ is satisfied.
* All quantities that have been empirically observed (or externally estimated from empirically observed data, and supplied to the model as "data") and are used in a likelihood component are denoted by a "hat", e.g., a survival term ($\hat{\phi}^{A \rightarrow B}_{y,j}$) and its logit-normal standard error ($\hat{\sigma}_{\phi^{A \rightarrow B}_{y,j}}$, which we use as a direct estimate of observation error variability).
* Stochastic processes are represented by $x_{y} \sim \mathcal{F}(\theta_{1}, \theta_{2})$, where $x_{y}$ is a random variable, $\mathcal{F}$ is some probability density (or mass) function ($\mathcal{N}$ is univariate normal, $\mathcal{MVN}$ is multivariate normal, $\mathcal{B}$ is binomial, and $\mathcal{M}$ is multinomial), and $\theta_{1}$ and $\theta_{2}$ are parameters of the probability distribution; generally $\theta_{1}$ is a location parameter specifying the expected value of $x_{y}$ and $\theta_{2}$ is a dispersion parameter specifying the variability of $x_{y}$ around the expected value.

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \rightarrow B}_{y,j}$ as an example.
We denote its random variability across years and populations as:

\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \rightarrow B}_{y,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \rightarrow B}_{y,1:n_j}\right), \Sigma_{\phi^{A \rightarrow B}} \right]
(\#eq:generic-random)
\end{equation}

where $\Sigma_{\phi^{A \rightarrow B}}$ is an $n_j \times n_j$ matrix constructed of population-specific variance terms $\left(\sigma_{\phi^{A \rightarrow B}_j}\right)^2$ and correlation terms specific to each pair of populations $j$ and $j^\prime$ $\rho_{\phi^{A \rightarrow B}_{j,j^{\prime}}}$:

\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \rightarrow B}_{j=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}_{j=1,j^{\prime}=2}} & \dots & \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}_{j=1,j^{\prime}=n_j}} \\
\sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}_{j=2,j^{\prime}=1}} & \left(\sigma_{\phi^{A \rightarrow B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \rho_{\phi^{A \rightarrow B}_{j=2,j^{\prime}=n_j}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=1}} \cdot \rho_{\phi^{A \rightarrow B}_{j=n_j,j^{\prime}=1}} & \sigma_{\phi^{A \rightarrow B}_{j=n_j}} \cdot \sigma_{\phi^{A \rightarrow B}_{j=2}} \cdot \rho_{\phi^{A \rightarrow B}_{j=n_j,j^{\prime}=2}} & \dots & \left(\sigma_{\phi^{A \rightarrow B}_{j=n_j}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma4by4)
\end{equation}

@@

\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \rightarrow B}_{o=1}}\right)^2 & \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \rho_{\phi^{A \rightarrow B}} \\
\sigma_{\phi^{A \rightarrow B}_{o=2}} \cdot \sigma_{\phi^{A \rightarrow B}_{o=1}} \cdot \rho_{\phi^{A \rightarrow B}} & \left(\sigma_{\phi^{A \rightarrow B}_{o=2}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma2by2)
\end{equation}

We estimated these covariance matrices by assigning their inverse ($\Sigma^{-1}$) a scaled Wishart prior distribution (Table \@ref(tab:param-table)).
Although the stochastic process requires the $\Sigma$ term, is is the component $\sigma$ and $\rho$ terms that we were interested in for inference.

@@

\begin{equation}
\dot{\phi}^{E \rightarrow P^{b}}_{y,j} = \frac{1}{\frac{1}{\alpha_j} + \frac{E_{y,j}}{\beta_j}}
(\#eq:get-phi-E-Pb-dot)
\end{equation}

@@

We modeled realized (i.e., with process noise) egg-to-parr survival as a multivariate logit-normal random variable around the expected value ($\dot{\phi}^{E \rightarrow P^{b}}_{y,j}$) from eq. \@ref(eq:get-phi-E-Pb-dot).
Rather than assume that inter-annual variability in egg-to-parr survival beyond that explained by density-dependence is completely random across years, we allowed the model to account for serial autocorrelation in egg survival anomalies via a lag-1 autoregressive process [AR(1); coefficient denoted $\kappa^{E \rightarrow P^{b}}_{j}$], wherein the white noise portion of the anomaly has covariance matrix $\Sigma_{\phi^{E \rightarrow P^{b}}}$:

\begin{equation}
\logit\left(\boldsymbol{\phi}^{E \rightarrow P^{b}}_{y,1:n_j}\right) \sim \MVNdist\left\{\logit\left(\boldsymbol{\dot{\phi}}^{E \rightarrow P^{b}}_{y,1:n_j}\right) + \boldsymbol{\kappa}^{E \rightarrow P^{b}}_{1:n_j} \cdot \left[\logit\left(\boldsymbol{\phi}^{E \rightarrow P^{b}}_{y-1,1:n_j}\right) - \logit\left(\boldsymbol{\dot{\phi}}^{E \rightarrow P^{b}}_{y-1,1:n_j}\right)\right], \Sigma_{\phi^{E \rightarrow P^{b}}} \right\}
(\#eq:phi-E-Pb-random)
\end{equation}

@@

\begin{equation}
P^{b}_{y,j} = \phi^{E \rightarrow P^{b}}_{y,j} \cdot E_{y,j}
(\#eq:get-Pb)
\end{equation}

@@

We modeled expected survival from the parr (age-1) to in-basin smolt (age-2) stage (termed "overwinter" survival and denoted by $\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}$) as a logit-linear function of parr mean length:

\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot L^{*P^{b}}_{y,j}
(\#eq:get-phi-Pa-Mb-dot)
\end{equation}

where $L^{*P^{b}}_{y,j}$ is parr mean length scaled and centered by subtracting the historical average observed parr mean length for population $j$ from $L^{P^{b}}_{y,j}$ and dividing by the historical observed standard deviation in mean length across years.
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{P^{a} \rightarrow M^{b}}}$ (assumed common among migratory strategies):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \rightarrow M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \rightarrow M^{b}}}\right]
(\#eq:phi-Pa-Mb-random)
\end{equation}

@@

\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \rightarrow M^{b}}_{y,i,j}
(\#eq:get-Mb-NOR)
\end{equation}

@@

\begin{equation}
\log\left(\dot{\Delta}^{L^{P^{b}} \rightarrow L^{M^{b}}}_{y,j}\right) = \theta_{0,j} + \theta_{1,j} \cdot L^{*P^{b}}_{y,j}
(\#eq:get-Delta-L-Pb-L-Mb-dot)
\end{equation}

and obtained the realized multiplicative change in mean length as being multivariate log-normally distributed around $\dot{\Delta}^{L^{P^{b}} \rightarrow L^{M^{b}}}_{y,j}$ with covariance matrix $\Sigma_{\Delta^{L^{P^{b}} \rightarrow L^{M^{b}}}}$:

\begin{equation}
\log\left(\boldsymbol{\Delta}^{L^{P^{b}} \rightarrow L^{M^{b}}}_{y,1:n_j}\right) \sim \MVNdist\left[\log\left(\boldsymbol{\dot{\Delta}}^{L^{P^{b} \rightarrow L^{M^{b}}}}_{y,1:n_j}\right), \Sigma_{\Delta^{L^{P^{b}} \rightarrow L^{M^{b}}}}\right]
(\#eq:Delta-L-Pb-L-Mb-random)
\end{equation}

@@

\begin{equation}
L^{M^{b}}_{y,j} = L^{P^{b}}_{y,j} \cdot \Delta^{L^{P^{b}} \rightarrow L^{M^{b}}}_{y,j}
(\#eq:get-L-Mb)
\end{equation}

@@

All in-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2 regardless of their migratory strategy ($i$).
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams\todo{Insert reference to a document describing dams} encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \rightarrow M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these quantities below), but we were able to model natural-origin survival separately from hatchery-origin because they are monitored separately (see Section \@ref(survival-data-sources)).

@@

\begin{equation}
\logit\left(\dot{\phi}^{M^{b} \rightarrow M^{a}}_{y,o=\mathrm{NOR},j}\right) = \tau_{0,j} + \tau_{1,j} \cdot L^{*M^{b}}_{y,j}
(\#eq:get-phi-Mb-Ma-dot)
\end{equation}

where $L^{*M^{b}}_{y,j}$ has been scaled and centered by subtracting the historical average observed smolt mean length for population $j$ from $L^{M^{b}}_{y,j}$ and dividing by the historical observed standard deviation (just as for $L^{*P^{b}}_{y,j}$ for standardized parr mean length).
For hatchery-origin smolt, we assumed this expected out-of-basin survival term was time-constant (i.e., all $\dot{\phi}^{M^{b} \rightarrow M^{a}}_{y,o=\mathrm{HOR},j}$ were equal within a population).
For each origin separately, we modeled realized out-of-basin migration survival as a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{M^{b} \rightarrow M^{a}}}$ (assumed common across origin types):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \rightarrow M^{a}}_{y,o,1:n_j}\right), \Sigma_{\phi^{M^{b} \rightarrow M^{a}}}\right]
(\#eq:phi-Mb-Ma-random)
\end{equation}

@@

\begin{equation}
M^{a}_{y,i,o,j} = M^{b}_{y,i,o,j} \cdot \phi^{M^{b} \rightarrow M^{a}}_{y,o,j}
(\#eq:get-Ma)
\end{equation}

>>>>

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \renewcommand*{\log}{\mathrm{log}\hspace{-0.20em}}
  - \renewcommand*{\max}{\mathrm{max}\hspace{-0.20em}}
  - \newcommand{\logit}{\mathrm{logit}\hspace{-0.20em}}
  - \newcommand{\ilogit}{\mathrm{logit}^{-1}\hspace{-0.20em}}
  - \newcommand{\MVNdist}{\mathcal{MVN}\hspace{-0.20em}}
  - \newcommand{\Ndist}{\mathcal{N}\hspace{-0.20em}}
  - \newcommand{\Bdist}{\mathcal{B}\hspace{-0.20em}}
  - \newcommand{\Mdist}{\mathcal{M}\hspace{-0.20em}}
  # - \newcommand{\fall}{\mathrm{fall}}
  # - \newcommand{\spring}{\mathrm{spring}}
  # - \newcommand{\NOR}{\mathrm{NOR}}
  # - \newcommand{\HOR}{\mathrm{HOR}}
---

@@

* For simplicity, we use the term "adult" to refer to any mature individual returning to spawn including age-3 "jacks".
* Life stage-specific states are defined by capital letters, e.g., parr ($P$), smolt ($M$), or spawners ($S$)
* Parameters (e.g., survival terms, non-survival transition probabilities, coefficients in relationships, etc.) are denoted by Greek symbols, e.g., $\phi$ for survival rates or $\psi$ for maturation rates.
* Subscripts denote the index of a quantity, e.g., $y$ denotes a specific year and $j$ a specific population (see Table \@ref(tab:index-table)).
* We use the convention $b$ to indicate "before event" and $a$ to indicate "after event" for a quantity that occurs at the same life stage. E.g., $M^{b}$ represents smolt abundance _before_ out-of-basin migration and $M^{a}$ represents smolt abundance _after_ out-of-basin migration. As such, superscripts are most often used syntactically, not mathematically. E.g., $M^{b}$ is not "$M$ raised to the power $b$"; if we must raise the quantity $M^{b}$ to the power $c$, it will be written $(M^{b})^c$.
* Survival terms to transition from state $A$ to state $B$ in year $y$ for population $j$ are denoted by $\phi^{A \to B}_{y,j}$.
* We denote the expected value of a stochastic process with a dot, e.g., $\dot{\phi}^{A \to B}_j$ -- this is the value expected in the absence of process variability.
* In some cases, we specify calculations performed on vectors rather than scalars (i.e., a scalar being a single element from a vector or higher-dimensional array). Vectors are denoted via boldface, and the dimension of the vector is denoted using 1:$n$ syntax. E.g., $\boldsymbol{\phi}^{A \to B}_{y,1:n_j}$ represents a vector of survival terms from state $A$ to state $B$ in year $y$, where each element of the vector represents the value for each population.
* Mathematical operations performed on two vectors $\boldsymbol{C}$ and $\boldsymbol{D}$ are performed in an element-wise fashion, and the resulting vector will have identical dimensions to both $\boldsymbol{C}$ and $\boldsymbol{D}$.
* In some cases, a calculation is performed differently for a subset of the possible index values. For example, to denote that $k$ should take on only the values of 1 and 3 (not 2), we would write, $k \in [1,3]$.
* $\log\left(x\right)$ represents the natural logarithm of $x$ and it is implied that the constraint $x > 0$ is satisfied.
* $\logit\left(p\right)$ represents the log odds of $p$ (i.e., $\log\left[p/(1-p)\right])$) and it is implied that the constraint $0 < p < 1$ is satisfied.
* All quantities that have been empirically observed (or externally estimated from empirically observed data, and supplied to the model as "data") and are used in a likelihood component are denoted by a "hat", e.g., a survival term ($\hat{\phi}^{A \to B}_{y,j}$) and its logit-normal standard error ($\hat{\sigma}_{\phi^{A \to B}_{y,j}}$, which we use as a direct estimate of observation error variability).
* Stochastic processes are represented by $x_{y} \sim \mathcal{F}(\theta_{1}, \theta_{2})$, where $x_{y}$ is a random variable, $\mathcal{F}$ is some probability density (or mass) function ($\mathcal{N}$ is univariate normal, $\mathcal{MVN}$ is multivariate normal, $\mathcal{B}$ is binomial, and $\mathcal{M}$ is multinomial), and $\theta_{1}$ and $\theta_{2}$ are parameters of the probability distribution; generally $\theta_{1}$ is a location parameter specifying the expected value of $x_{y}$ and $\theta_{2}$ is a dispersion parameter specifying the variability of $x_{y}$ around the expected value.

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \to B}_{y,j}$ as an example.
We denote its random variability across years and populations as:

\begin{equation}
\logit\left(\boldsymbol{\phi}^{A \to B}_{y,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{A \to B}_{y,1:n_j}\right), \Sigma_{\phi^{A \to B}} \right]
(\#eq:generic-random)
\end{equation}

where $\Sigma_{\phi^{A \to B}}$ is an $n_j \times n_j$ matrix constructed of population-specific variance terms $\left(\sigma_{\phi^{A \to B}_j}\right)^2$ and correlation terms specific to each pair of populations $j$ and $j^\prime$ $\rho_{\phi^{A \to B}_{j,j^{\prime}}}$:

\begin{equation}
\left[
\begin{array}{cccc}
\left(\sigma_{\phi^{A \to B}_{j=1}}\right)^2 & \sigma_{\phi^{A \to B}_{j=1}} \cdot \sigma_{\phi^{A \to B}_{j=2}} \cdot \rho_{\phi^{A \to B}_{j=1,j^{\prime}=2}} & \dots & \sigma_{\phi^{A \to B}_{j=1}} \cdot \sigma_{\phi^{A \to B}_{j=n_j}} \cdot \rho_{\phi^{A \to B}_{j=1,j^{\prime}=n_j}} \\
\sigma_{\phi^{A \to B}_{j=2}} \cdot \sigma_{\phi^{A \to B}_{j=1}} \cdot \rho_{\phi^{A \to B}_{j=2,j^{\prime}=1}} & \left(\sigma_{\phi^{A \to B}_{j=2}}\right)^2 & \dots & \sigma_{\phi^{A \to B}_{j=2}} \cdot \sigma_{\phi^{A \to B}_{j=n_j}} \cdot \rho_{\phi^{A \to B}_{j=2,j^{\prime}=n_j}} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{\phi^{A \to B}_{j=n_j}} \cdot \sigma_{\phi^{A \to B}_{j=1}} \cdot \rho_{\phi^{A \to B}_{j=n_j,j^{\prime}=1}} & \sigma_{\phi^{A \to B}_{j=n_j}} \cdot \sigma_{\phi^{A \to B}_{j=2}} \cdot \rho_{\phi^{A \to B}_{j=n_j,j^{\prime}=2}} & \dots & \left(\sigma_{\phi^{A \to B}_{j=n_j}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma4by4)
\end{equation}

@@

\begin{equation}
\left[
\begin{array}{cc}
\left(\sigma_{\phi^{A \to B}_{o=1}}\right)^2 & \sigma_{\phi^{A \to B}_{o=1}} \cdot \sigma_{\phi^{A \to B}_{o=2}} \cdot \rho_{\phi^{A \to B}} \\
\sigma_{\phi^{A \to B}_{o=2}} \cdot \sigma_{\phi^{A \to B}_{o=1}} \cdot \rho_{\phi^{A \to B}} & \left(\sigma_{\phi^{A \to B}_{o=2}}\right)^2
\end{array}
\right]
(\#eq:generic-Sigma2by2)
\end{equation}

We estimated these covariance matrices by assigning their inverse ($\Sigma^{-1}$) a scaled Wishart prior distribution (Table \@ref(tab:param-table)).
Although the stochastic process requires the $\Sigma$ term, it is the component $\sigma$ and $\rho$ terms that we were interested in for inference.

@@

\begin{equation}
\dot{\phi}^{E \to P^{b}}_{y,j} = \frac{1}{\frac{1}{\alpha_j} + \frac{E_{y,j}}{\beta_j}}
(\#eq:get-phi-E-Pb-dot)
\end{equation}

@@

We modeled realized (i.e., with process noise) egg-to-parr survival as a multivariate logit-normal random variable around the expected value ($\dot{\phi}^{E \to P^{b}}_{y,j}$) from eq. \@ref(eq:get-phi-E-Pb-dot).
Rather than assume that inter-annual variability in egg-to-parr survival beyond that explained by density-dependence is completely random across years, we allowed the model to account for serial autocorrelation in egg survival anomalies via a lag-1 autoregressive process [AR(1); coefficient denoted $\kappa^{E \to P^{b}}_{j}$], wherein the white noise portion of the anomaly has covariance matrix $\Sigma_{\phi^{E \to P^{b}}}$:

\begin{equation}
\logit\left(\boldsymbol{\phi}^{E \to P^{b}}_{y,1:n_j}\right) \sim \MVNdist\left\{\logit\left(\boldsymbol{\dot{\phi}}^{E \to P^{b}}_{y,1:n_j}\right) + \boldsymbol{\kappa}^{E \to P^{b}}_{1:n_j} \cdot \left[\logit\left(\boldsymbol{\phi}^{E \to P^{b}}_{y-1,1:n_j}\right) - \logit\left(\boldsymbol{\dot{\phi}}^{E \to P^{b}}_{y-1,1:n_j}\right)\right], \Sigma_{\phi^{E \to P^{b}}} \right\}
(\#eq:phi-E-Pb-random)
\end{equation}

@@

\begin{equation}
P^{b}_{y,j} = \phi^{E \to P^{b}}_{y,j} \cdot E_{y,j}
(\#eq:get-Pb)
\end{equation}

@@

We modeled expected survival from the parr (age-1) to in-basin smolt (age-2) stage (termed "overwinter" survival and denoted by $\dot{\phi}^{P^{a} \to M^{b}}_{y,i,j}$) as a logit-linear function of parr mean length:

\begin{equation}
\logit\left(\dot{\phi}^{P^{a} \to M^{b}}_{y,i,j}\right) = \gamma_{0,i,j} + \gamma_{1,j} \cdot L^{*P^{b}}_{y,j}
(\#eq:get-phi-Pa-Mb-dot)
\end{equation}

where $L^{*P^{b}}_{y,j}$ is parr mean length scaled and centered by subtracting the historical average observed parr mean length for population $j$ from $L^{P^{b}}_{y,j}$ and dividing by the historical observed standard deviation in mean length across years.
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{P^{a} \to M^{b}}}$ (assumed common among migratory strategies):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{P^{a} \to M^{b}}_{y,i,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{P^{a} \to M^{b}}_{y,i,1:n_j}\right), \Sigma_{\phi^{P^{a} \to M^{b}}}\right]
(\#eq:phi-Pa-Mb-random)
\end{equation}

@@

\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \to M^{b}}_{y,i,j}
(\#eq:get-Mb-NOR)
\end{equation}

@@

\begin{equation}
\log\left(\dot{\Delta}^{L^{P^{b}} \to L^{M^{b}}}_{y,j}\right) = \theta_{0,j} + \theta_{1,j} \cdot L^{*P^{b}}_{y,j}
(\#eq:get-Delta-L-Pb-L-Mb-dot)
\end{equation}

and obtained the realized multiplicative change in mean length as being multivariate log-normally distributed around $\dot{\Delta}^{L^{P^{b}} \to L^{M^{b}}}_{y,j}$ with covariance matrix $\Sigma_{\Delta^{L^{P^{b}} \to L^{M^{b}}}}$:

\begin{equation}
\log\left(\boldsymbol{\Delta}^{L^{P^{b}} \to L^{M^{b}}}_{y,1:n_j}\right) \sim \MVNdist\left[\log\left(\boldsymbol{\dot{\Delta}}^{L^{P^{b} \to L^{M^{b}}}}_{y,1:n_j}\right), \Sigma_{\Delta^{L^{P^{b}} \to L^{M^{b}}}}\right]
(\#eq:Delta-L-Pb-L-Mb-random)
\end{equation}

@@

\begin{equation}
L^{M^{b}}_{y,j} = L^{P^{b}}_{y,j} \cdot \Delta^{L^{P^{b}} \to L^{M^{b}}}_{y,j}
(\#eq:get-L-Mb)
\end{equation}

@@

All in-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2 regardless of their migratory strategy ($i$).
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams\todo{Insert reference to a document describing dams} encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \to M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these quantities below), but we were able to model natural-origin survival separately from hatchery-origin because they are monitored separately (see Section \@ref(survival-data-sources)).

@@

\begin{equation}
\logit\left(\dot{\phi}^{M^{b} \to M^{a}}_{y,o=\mathrm{NOR},j}\right) = \tau_{0,j} + \tau_{1,j} \cdot L^{*M^{b}}_{y,j}
(\#eq:get-phi-Mb-Ma-dot)
\end{equation}

where $L^{*M^{b}}_{y,j}$ has been scaled and centered by subtracting the historical average observed smolt mean length for population $j$ from $L^{M^{b}}_{y,j}$ and dividing by the historical observed standard deviation (just as for $L^{*P^{b}}_{y,j}$ for standardized parr mean length).
For hatchery-origin smolt, we assumed this expected out-of-basin survival term was time-constant (i.e., all $\dot{\phi}^{M^{b} \to M^{a}}_{y,o=\mathrm{HOR},j}$ were equal within a population).
For each origin separately, we modeled realized out-of-basin migration survival as a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{M^{b} \to M^{a}}}$ (assumed common across origin types):

\begin{equation}
\logit\left(\boldsymbol{\phi}^{M^{b} \to M^{a}}_{y,o,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\phi}}^{M^{b} \to M^{a}}_{y,o,1:n_j}\right), \Sigma_{\phi^{M^{b} \to M^{a}}}\right]
(\#eq:phi-Mb-Ma-random)
\end{equation}

@@

\begin{equation}
M^{a}_{y,i,o,j} = M^{b}_{y,i,o,j} \cdot \phi^{M^{b} \to M^{a}}_{y,o,j}
(\#eq:get-Ma)
\end{equation}

We modeled survival during the downstream migration through the hydrosystem on the Snake and Columbia rivers as a single survival rate, specific to each year and origin but shared among populations ($\phi^{M^{a} \to O^{0}}_{y,o}$).
We assumed these survival rates were multivariate logit-normal random variables around a time-constant expected value ($\dot{\phi}^{M^{a} \to O^{0}}_{o}$) with covariance matrix $\Sigma_{\phi^{M^{a} \to O^{0}}}$:
